---
title: "Binary Multiple Predictors"
#output: html_notebook
---

Running predictions with multiple predictors based on a dag. This means using **ALL** the top 10 predictors in one model, along with size (log_Mass). No Interactions. 

# Preparation  

### Load packages and functions  
```{r message = FALSE, warning=FALSE, include = FALSE}

pacman::p_load(tidyverse,  brms, cmdstanr, kableExtra, here, ape, tidybayes, patchwork, glue, ggthemes)

options(brms.backend = "cmdstanr")

here::i_am("README.md")
here()
```



### Load data  

All data loading and wrangling in `Code/Data_Wrangle.Rmd`
```{r}
dat <- read_csv(here("Data", "Clean_Data.csv"))

#species in dataset
spp_obs <- dat$Taxon_Upham_style %>% unique()

# Organizing complete measurements and those with missing values.
# they require different models
var_complete <-  read_rds(here("Data", "var_complete.Rds"))
var_missing <-  read_rds(here("Data", "var_missing.Rds"))
var_complete_linear <-  read_rds(here("Data", "var_complete_linear.Rds"))
var_missing_linear <-  read_rds(here("Data", "var_missing_linear.Rds"))
var_complete_ratio <-  read_rds(here("Data", "var_complete_ratio.Rds"))
var_missing_ratio <-  read_rds(here("Data", "var_missing_ratio.Rds"))
var_accurate <- read_rds(here("Data", "var_accurate.Rds"))

#Color palette
pal = c('#62361B', '#f48a1a', '#e6ab02', '#66a61e', '#31c1bb', '#0d57ff', '#5e1ea6', '#f9581c' , '#f384f8')


# Load Tree & Get Cor Matrix

tr <- ape::read.nexus(
  here("Data","MCC_Upham_all_mammals.nex"))
tr <- ape::keep.tip(tr, spp_obs)
A <- ape::vcv.phylo(tr, corr = TRUE)
```

## Multiple Regression Model


Phylo Trick:
```{r}
# 1. Create indicator
dat <- dat %>%
  mutate(
    Loc_bin_obs = ifelse(is.na(Loc_bin), -1L, as.integer(Loc_bin)),
    y_for_brms = ifelse(is.na(Loc_bin), 0L, as.integer(Loc_bin))
  )

# 2. Define custom family
bernoulli_missing <- custom_family(
  name = "bernoulli_missing",
  dpars = "mu",
  links = "logit",
  lb = 0, ub = 1,
  type = "int",
  vars = "vint1[n]"
)

stan_funs <- "
  real bernoulli_missing_lpmf(int y_placeholder, real mu, int y_obs) {
    if (y_obs >= 0) {
      return bernoulli_lpmf(y_obs | mu);
    } else {
      return 0;
    }
  }
"
stanvars <- stanvar(scode = stan_funs, block = "functions")
```

Here are the accurate measurements:
"HPI"   "PES"   "MANUS" "Ppl"   "Hl"    "OLI"   "Rl"    "Uol"   "Fl"    "Ul"   
With complete data are:
Fl, Hl, HPI, Uol
With missing data are:
PES, MANUS, Ppl, Ul, OLI, Rl

#### Model Objects
```{r}
bin_mod <-  bf(y_for_brms | vint(Loc_bin_obs) ~ Fl + Hl + HPI + Uol + mi(PES) + mi(MANUS) + mi(Ppl) + mi(Ul)
               + mi(OLI) + mi(Rl) + mi(log_Mass) + (1|Genus_species) 
               + (1|gr(Taxon_Upham_style,cov = A))) + bernoulli_missing
  
PES_mod <- bf(PES| mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl + (1|Genus_species) + 
                (1|gr(Taxon_Upham_style, cov = A))) + gaussian()

MANUS_mod <- bf(MANUS| mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl + (1|Genus_species) + 
                (1|gr(Taxon_Upham_style, cov = A))) + gaussian()

Ppl_mod <- bf(Ppl| mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl + (1|Genus_species) + 
                (1|gr(Taxon_Upham_style, cov = A))) + gaussian()

Ul_mod <- bf(Ul| mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl  + (1|Genus_species) + 
                (1|gr(Taxon_Upham_style, cov = A))) + gaussian()

OLI_mod <- bf(OLI| mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl  + (1|Genus_species) + 
                (1|gr(Taxon_Upham_style, cov = A))) + gaussian()

Rl_mod <- bf(Rl| mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl  + (1|Genus_species) + 
                (1|gr(Taxon_Upham_style, cov = A))) + gaussian()

mass_mod <- bf(log_Mass| mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl + (1|Genus_species) + 
                (1|gr(Taxon_Upham_style, cov = A))) + gaussian()

m_form <- bin_mod + PES_mod + MANUS_mod + Ppl_mod + Ul_mod + OLI_mod + Rl_mod + mass_mod + set_rescor(FALSE)
  
pri <- get_prior(m_form,
  data2 = list(A = A),
           data = dat) %>% 
  mutate(
    prior = case_when(
      class == "b" & coef == ""  ~ "normal(0, 2)",
      class == "sd" & coef == ""  ~ "normal(0, 0.3)",
      class == "Intercept" ~ "normal(0, 2)",
      class == "sigma" ~ "normal(0, 0.5)",
      TRUE ~ prior
      ))


multmod <- brm(m_form,
                    data=dat,
                    refresh = 0,
                    cores = 4,
                    data2 = list(A = A),
                    prior = pri,
                    seed = 145,
                    control=list(max_treedepth=15),
                    silent = 2,
                    stanvars = stanvars
               )
```

## Save Models
```{r}
multmod %>% write_rds(here("Data","B_mod_multi.Rds"))
```

# Predictions  

```{r}
mice <- c("Podomys_floridanus", "Ochrotomys_nuttalli","Peromyscus_gossypinus", "Peromyscus_polionotus")

model_dat <- dat %>% filter(Genus_species %in% c(mice)) |>
  mutate(
    y_for_brms = 0L,
    Loc_bin_obs = -1L
  )
```


```{r}
posterior_epred_bernoulli_missing <- function(prep) {
  mu <- brms::get_dpar(prep, "mu")
  mu
}

pred_df_ext_full <- multmod %>% 
    add_epred_draws(newdata = model_dat, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE, 
                    re_formula = NULL) %>% 
    as.data.frame() %>% 
    select(Genus_species, Prob_of_Climbing, catalog) %>% 
  mutate(method = "Full")
    

pred_df_ext_fos <- multmod %>% 
    add_epred_draws(newdata = model_dat, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    #allow_new_levels = TRUE, 
                    re_formula = NA) %>% 
    as.data.frame() %>% 
    select(Genus_species, Prob_of_Climbing, catalog) %>% 
  mutate(method = "NoPhylo")
```

# Average Over Individuals
```{r}
pred_species <-multmod %>%
    add_epred_draws(newdata = model_dat, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    #allow_new_levels = TRUE, 
                    re_formula = NULL) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop")
```

# Plot

```{r}
pred_df_ext_full %>%
  group_by(catalog, Genus_species) %>%
  mutate(mean_prob = mean(Prob_of_Climbing, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(catalog = reorder(catalog, mean_prob),
         catalog = reorder(catalog, as.numeric(factor(Genus_species)))) %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = catalog,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89)) +
  ylim(0, 1) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Average Over Individuals
```{r}
pred_species <-multmod %>%
    add_epred_draws(newdata = model_dat, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE, 
                    re_formula = NULL) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop")

p1 <- pred_species %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = Genus_species,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89)) +
  ylim(0, 1) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

pred_species <-multmod %>%
    add_epred_draws(newdata = model_dat, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE, 
                    re_formula = NA) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop")

p2 <- pred_species %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = Genus_species,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89)) +
  ylim(0, 1) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

pred_species <-multmod %>%
    add_epred_draws(newdata = model_dat, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    #allow_new_levels = TRUE, 
                    re_formula = NULL) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop")

p3 <- pred_species %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = Genus_species,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89)) +
  ylim(0, 1) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

pred_species <-multmod %>%
    add_epred_draws(newdata = model_dat, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    #allow_new_levels = TRUE, 
                    re_formula = NA) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop")

p4 <- pred_species %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = Genus_species,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89)) +
  ylim(0, 1) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

(p1 + p2) / (p3 + p4) + plot_layout(guides = "collect")
```

```{r}
hplot <- function(groups, method_choice){
  pred_all %>% 
    filter(Genus_species %in% {{groups}},
           method == {{method_choice}}) %>% 
    ggplot(aes(y = Prob_of_Climbing, 
               x = catalog, 
               color = Genus_species)) +
    geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
    stat_pointinterval(.width = c(0.5, 0.68268), 
                       position = position_dodge(width = -.6)) + 
    ylim(0,1) +
    labs(x = "", y = "Probability of Climbing") +
    scale_color_manual(values = pal) + 
    facet_wrap(~Genus_species, scales = "free_x") +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
}
```



```{r}

pred_all <- bind_rows(pred_df_ext_full, pred_df_ext_fos, pred_df) %>% #filter(var %in% var_acc) %>% 
  mutate(draw = rep(1:4000, times = nrow(.)/4000)) %>% 
  #THIS IS AVERAGING PER SPECIES
  group_by(substrate, method, Genus_species, draw) %>% 
  summarize(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE)) %>% ungroup() %>% mutate(var = "combined") %>% 
  write_csv(here("Data", "Binary_Mult_Preds.csv"))

```


# Predictions 2

```{r}

mice <- c("Podomys_floridanus", "Ochrotomys_nuttalli","Peromyscus_gossypinus", "Peromyscus_polionotus", "Cynocephalus_volans")

posterior_epred_bernoulli_missing <- function(prep) {
  mu <- brms::get_dpar(prep, "mu")
  mu
}

model_dat <- dat %>% filter(Genus_species %in% c(mice)) |>
  mutate(
    y_for_brms = 0L,
    Loc_bin_obs = -1L
  )

# Full model (both REs)
pred_full <- multmod %>% 
    add_epred_draws(newdata = model_dat,
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE,
                    re_formula = NULL) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop") %>%
  mutate(method = "Full")

# Species RE only (no phylo)
pred_species_only <- multmod %>% 
    add_epred_draws(newdata = model_dat,
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE,
                    re_formula = ~(1|Genus_species)) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop") %>%
  mutate(method = "Species_RE")

# Phylo only (no species RE)
pred_phylo_only <- multmod %>% 
    add_epred_draws(newdata = model_dat,
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE,
                    re_formula = ~(1|gr(Taxon_Upham_style, cov = A))) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop") %>%
  mutate(method = "Phylo_only")

# Fixed effects only
pred_fixef <- multmod %>% 
    add_epred_draws(newdata = model_dat,
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE,
                    re_formula = NA) %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop") %>%
  mutate(method = "Fixed_only")

# Combine
pred_all <- bind_rows(pred_full, pred_species_only, pred_phylo_only, pred_fixef) %>%
  mutate(method = factor(method, levels = c("Full", "Species_RE", "Phylo_only", "Fixed_only")))

# Plot
pred_all %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = method,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89),
                     position = position_dodge(width = 0.4)) +
  ylim(0, 1) +
  facet_wrap(~Genus_species) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


```{r}
posterior_epred_bernoulli_missing <- function(prep) {
  mu <- brms::get_dpar(prep, "mu")
  mu
}
```



# PRediction 3 Imputation
```{r}
model_dat_imp <- model_dat %>%
  group_by(Genus_species) %>%
  mutate(across(c(PES, MANUS, Ppl, Ul, OLI, Rl, log_Mass),
                ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()


posterior_epred_bernoulli_missing <- function(prep) {
  mu <- brms::get_dpar(prep, "mu")
  mu
}

pred_df_ext_full <- multmod %>% 
    add_epred_draws(newdata = model_dat_imp, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE, 
                    re_formula = NULL) %>% 
    as.data.frame() %>% 
    select(Genus_species, Prob_of_Climbing, catalog) %>% 
  mutate(method = "Full")

pred_df_ext_full %>%
  group_by(catalog, Genus_species) %>%
  mutate(mean_prob = mean(Prob_of_Climbing, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(catalog = reorder(catalog, mean_prob),
         catalog = reorder(catalog, as.numeric(factor(Genus_species)))) %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = catalog,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89)) +
  ylim(0, 1) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Prediction 4 - complete
```{r}
model_dat_complete <- model_dat %>%
  filter(!is.na(PES) | TRUE) %>%  # mi() variables are fine
  # actually drop specimens missing complete-data predictors
  # plus any with NA in mi() variables that weren't in training data
  anti_join(
    pred_df_ext_full %>%
      group_by(catalog) %>%
      summarise(prop_na = mean(is.na(Prob_of_Climbing))) %>%
      filter(prop_na > 0),
    by = "catalog"
  )


posterior_epred_bernoulli_missing <- function(prep) {
  mu <- brms::get_dpar(prep, "mu")
  mu
}

pred_df_ext_full <- multmod %>% 
    add_epred_draws(newdata = model_dat_complete, 
                    resp = "yforbrms",
                    value = "Prob_of_Climbing",
                    allow_new_levels = TRUE, 
                    re_formula = NULL) %>% 
    as.data.frame() %>% 
    select(Genus_species, Prob_of_Climbing, catalog) %>% 
  mutate(method = "Full")

pred_df_ext_full %>%
  group_by(catalog, Genus_species) %>%
  mutate(mean_prob = mean(Prob_of_Climbing, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(catalog = reorder(catalog, mean_prob),
         catalog = reorder(catalog, as.numeric(factor(Genus_species)))) %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = catalog,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89)) +
  ylim(0, 1) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Predicting missing values per measurement per species


```{r}
tictoc::tic()
plan(multisession, workers = 8)

chunk_size <- 250

thin_draws <- seq(1, 4000, by = 4)  # every 4th draw = 1000 total
chunks <- split(thin_draws, ceiling(seq_along(thin_draws) / chunk_size))

for (chunk_i in seq_along(chunks)) {
  chunk_draws <- chunks[[chunk_i]]
  message(glue("Starting chunk {chunk_i} of {length(chunks)}"))
  
  chunk_results <- future_map(chunk_draws, ~{
    d <- .x
    
    imputed_d <- ymi_long %>%
      filter(.draw == d) %>%
      select(row_idx, variable, imputed) %>%
      pivot_wider(names_from = variable, values_from = imputed) %>%
      bind_rows(empty_imp)
    
    newdata_d <- model_dat_indexed %>%
      left_join(imputed_d, by = "row_idx", suffix = c("", "_imp")) %>%
      mutate(
        PES      = coalesce(PES, PES_imp),
        MANUS    = coalesce(MANUS, MANUS_imp),
        Ppl      = coalesce(Ppl, Ppl_imp),
        Ul       = coalesce(Ul, Ul_imp),
        OLI      = coalesce(OLI, OLI_imp),
        Rl       = coalesce(Rl, Rl_imp),
        log_Mass = coalesce(log_Mass, log_Mass_imp)
      ) %>%
      select(-ends_with("_imp"), -row_idx)
    
    multmod %>%
  add_epred_draws(newdata = newdata_d,
                  resp = "yforbrms",
                  value = "Prob_of_Climbing",
                  allow_new_levels = TRUE,
                  re_formula = NULL,
                  draw_ids = d) %>%
  as.data.frame() %>%
  select(Genus_species, catalog, Prob_of_Climbing) %>%
  mutate(.draw = d)  # tag with the actual draw number
  },
  .options = furrr_options(
    seed = TRUE,
    globals = c("ymi_long", "model_dat_indexed", "empty_imp", "multmod",
                "posterior_epred_bernoulli_missing"),
    packages = c("dplyr", "tidyr", "brms", "tidybayes")
  ))
  
  saveRDS(chunk_results, here("Data", glue("pred_chunk_{chunk_i}.rds")))
  message(glue("Chunk {chunk_i} saved"))
}

plan(sequential)

tictoc::toc()
# After all chunks done, reassemble
pred_proper <- map(seq_along(chunks), ~{
  readRDS(here("Data", glue("pred_chunk_{.x}.rds")))
}) %>%
  flatten() %>%
  bind_rows()
```
Now average and plotting the proper predictions

```{r}
pred_proper_species <- pred_proper %>%
  group_by(Genus_species, .draw) %>%
  summarise(Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE), .groups = "drop")

pred_proper_species %>%
  ggplot(aes(y = Prob_of_Climbing,
             x = Genus_species,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.89)) +
  ylim(0, 1) +
  labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```




# Whole Shebang  


### function
```{r}

hplot <- function(groups, method){
  pred_all %>% filter(Genus_species %in% {{groups}},
                         method == {{method}}) %>% 
  ggplot(aes(y = Prob_of_Climbing, 
             x = Genus_species, 
             #y = var,
             color = Genus_species)) +
  geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.5) +
  stat_pointinterval(.width = c(0.5, 0.68268), 
                     #linewidth = c(0.75, 0.5), 
                     position = position_dodge(width = -.6)) + ylim(0,1) +
    labs(x = "", y = "Probability of Climbing") +
  scale_color_manual(values = pal) + theme_bw() + 
  theme(axis.text.x=element_text(angle=45, hjust=1))
}

```

###  big df



### Big Plot
```{r}
pl5 <- hplot(old, "Fossil") + 
  ggtitle("Mesozoic Mammal Predictions") 

pl6 <- hplot(multi, "Fossil") + 
  ggtitle("Multituberculate Predicitons") 

pl7 <- hplot(primate, "Fossil") + 
  ggtitle("Primate Predicitons") 


t3 <- pl5 / pl6 /pl7

t3

ggsave(here("Plots", "Predict_Bin_Mult_Fossil.pdf"), height = 8, width = 13)

```


```{r}
pl1 <- hplot(extant_climb, "Full") + 
  ggtitle("Extant Climb Full Predictions") 

pl2 <- hplot(extant_climb, "Fossil") + 
  ggtitle("Extant Climb Fossil") 

pl3 <- hplot(extant_ground, "Full") + 
  ggtitle("Extant Ground Full Predictions") 

pl4 <- hplot(extant_ground, "Fossil") + 
  ggtitle("Extant Ground Fossil") 


t1 <- pl1 / pl2 + plot_layout(guides = "collect")

t2 <- pl3 / pl4 + plot_layout(guides = "collect")

wrap_plots(t1, t2, ncol = 1)

ggsave(here("Plots", "Predict_Bin_Mult_Extant.pdf"), height = 10, width = 8)

```

```{r}
pl2 / pl4 + plot_layout(guides = "collect")

ggsave(here("Plots", "Predict_Bin_Mult_Extant2.pdf"), height = 6, width = 13)
```



```{r}
wrap_plots(t1, t2, t3, ncol = 1, heights = c(2/7, 2/7, 3/7))

ggsave(here("Plots", "Predict_Bin_Mult_All.pdf"), height = 16, width = 13)
```
