---
title: "Binary Models"
#output: html_notebook
---

# Binary Models Script

Predicting binary climbing classifications (tree vs. ground) using postcranial measurements. Models also incorporate intraspecific variation and phylogenetic variation.

After extensive trials, we have decided on:
  - log-shape ratio models will not include a size covariate
  - functional index models will not include a size covariate. Why? Testing showed that it didn't affect effect sizes or predictions. Go into detail(s) in Sup if needed. 
  - Linear measurements will include size, both gm and log-mass (because log-mass is needed for fossil predictions unless we want to do some big missing data models?)

## Preparation  

#### Load packages and functions  
```{r message = FALSE, warning=FALSE, include = FALSE}

pacman::p_load(tidyverse,  brms, cmdstanr, kableExtra, here, ape, tidybayes, patchwork, glue)

options(brms.backend = "cmdstanr")

scale2 <- function(x, na.rm = TRUE) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm)

here::i_am("README.md")
here()
```

#### Load data  
All data loading and wrangling in `Code/Data_Load.Rmd`
```{r}
dat <- read_csv(here("Data", "Clean_Data.csv"))

#species in dataset
spp_obs <- dat$Taxon_Upham_style %>% unique()

# Organizing complete measurements and those with missing values.
# they require different models
var_complete <-  read_rds(here("Data", "var_complete.Rds"))
var_missing <-  read_rds(here("Data", "var_missing.Rds"))
var_complete_linear <-  read_rds(here("Data", "var_complete_linear.Rds"))
var_missing_linear <-  read_rds(here("Data", "var_missing_linear.Rds"))
var_complete_ratio <-  read_rds(here("Data", "var_complete_ratio.Rds"))
var_missing_ratio <-  read_rds(here("Data", "var_missing_ratio.Rds"))
```

#### Load Tree & Get Cor Matrix

```{r}
tr <- ape::read.nexus(
  here("Data","MCC_Upham_all_mammals.nex"))
tr <- ape::keep.tip(tr, spp_obs)
A <- ape::vcv.phylo(tr, corr = TRUE)
```

# Fit Models

## Linear Measurements

### Log-Mass & Variable Missing Missing

Update cannot be used with multivatiate models. Prior must be called in the loop

#### Run Loop

Create list for models to go into
```{r}
list_b_lm <- vector(mode ="list")
```


*Tricky new Phylo Model*   

This is an addition where I can run the model, then remove the specimens that need predictions from the liklihood estimate. This way I can use the phylogeny for predictions. 
```{r}
# 1. Create indicator
dat <- dat %>%
  mutate(
    Loc_bin_obs = ifelse(is.na(Loc_bin), -1L, as.integer(Loc_bin)),
    y_for_brms = ifelse(is.na(Loc_bin), 0L, as.integer(Loc_bin))
  )

# 2. Define custom family
bernoulli_missing <- custom_family(
  name = "bernoulli_missing",
  dpars = "mu",
  links = "logit",
  lb = 0, ub = 1,
  type = "int",
  vars = "vint1[n]"
)

stan_funs <- "
  real bernoulli_missing_lpmf(int y_placeholder, real mu, int y_obs) {
    if (y_obs >= 0) {
      return bernoulli_lpmf(y_obs | mu);
    } else {
      return 0;
    }
  }
"
stanvars <- stanvar(scode = stan_funs, block = "functions")
```


```{r}
for(i in var_missing_linear){
  bin_mod <-  bf(paste0("y_for_brms | vint(Loc_bin_obs) ~ mi(", i, ")+mi(log_Mass)+(1|Genus_species)+(1|gr(Taxon_Upham_style,cov = A))")) + bernoulli_missing
  imp_mod <- bf(paste0(i,"| mi() ~ Fl + Fdw +
       Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl + 
       (1|gr(Taxon_Upham_style, cov = A))")) + gaussian()
  mass_mod <- bf(paste0("log_Mass| mi() ~ Fl + Fdw +
       Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl + 
       (1|gr(Taxon_Upham_style, cov = A))")) + gaussian()
  m_form <- bin_mod + imp_mod + mass_mod + set_rescor(FALSE)
  
  pri <- get_prior(m_form,
  data2 = list(A = A),
           data = dat) %>% 
  mutate(
    prior = case_when(
      class == "b" & coef == ""  ~ "normal(0, 2)",
      class == "sd" & coef == ""  ~ "normal(0, 0.3)",
      class == "Intercept" ~ "normal(0, 2)",
      class == "sigma" ~ "normal(0, 0.5)",
      TRUE ~ prior
      ))

  list_b_lm[[i]]<- brm(m_form,
                         data=dat,
                         refresh = 0,
                         cores = 4,
                         data2 = list(A = A),
                         prior = pri,
                         seed = 145,
                         silent = 2,
                         stanvars = stanvars
                         ) 
}
```

## Linear only log_Mass missing

```{r}
for(i in var_complete_linear){
  bin_mod <-  bf(paste0("y_for_brms | vint(Loc_bin_obs) ~ ",i, "+mi(log_Mass)+(1|Genus_species)+(1|gr(Taxon_Upham_style,cov = A))")) + bernoulli_missing
  mass_mod <- bf(paste0("log_Mass| mi() ~ Fl + Fdw +
       Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl + 
       (1|gr(Taxon_Upham_style, cov = A))")) + gaussian()
  m_form <- bin_mod +  mass_mod + set_rescor(FALSE)
  
  pri <- get_prior(m_form,
  data2 = list(A = A),
           data = dat) %>% 
  mutate(
    prior = case_when(
      class == "b" & coef == ""  ~ "normal(0, 2)",
      class == "sd" & coef == ""  ~ "normal(0, 0.3)",
      class == "Intercept" ~ "normal(0, 2)",
      class == "sigma" ~ "normal(0, 0.5)",
      TRUE ~ prior
      ))

  list_b_lm[[i]]<- brm(m_form,
                         data=dat,
                         refresh = 0,
                         cores = 4,
                         data2 = list(A = A),
                         prior = pri,
                         seed = 145,
                         silent = 2,
                         stanvars = stanvars
                         ) 
}
```


## Ratios


### Ratios Missing

#### Run Loop

```{r}
for(i in var_missing_ratio){
  bin_mod <-  bf(paste0("y_for_brms | vint(Loc_bin_obs) ~ mi(", i, ")+(1|Genus_species)+(1|gr(Taxon_Upham_style,cov = A))")) + bernoulli_missing
  imp_mod <- bf(paste0(i,"| mi() ~ Fl + Fdw +
       Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl + 
       (1|gr(Taxon_Upham_style, cov = A))")) + gaussian()
  m_form <- bin_mod + imp_mod + mass_mod + set_rescor(FALSE)
  
  pri <- get_prior(m_form,
  data2 = list(A = A),
           data = dat) %>% 
  mutate(
    prior = case_when(
      class == "b" & coef == ""  ~ "normal(0, 2)",
      class == "sd" & coef == ""  ~ "normal(0, 0.3)",
      class == "Intercept" ~ "normal(0, 2)",
      class == "sigma" ~ "normal(0, 0.5)",
      TRUE ~ prior
      ))

  list_b_lm[[i]]<- brm(m_form,
                         data=dat,
                         refresh = 0,
                         cores = 4,
                         data2 = list(A = A),
                         prior = pri,
                         seed = 145,
                         silent = 2,
                         stanvars = stanvars
                         ) 
}
```


### Ratios Complete

#### Prelim Model 

```{r}

list_b_ratio <- vector(mode ="list")
# prior

p1 = c(prior(normal(0, 5), class = Intercept),
       prior(normal(0, 2), class = b),
       prior(normal(0, 0.3), class = sd))

#initial fit
mm <- brm(Loc_bin ~ Fl  + (1 | Genus_species) +  (1|gr(Taxon_Upham_style, cov = A)),
          data2 = list(A = A),
          family = bernoulli(),
          data = dat, refresh = 0, cores = 4,
          prior = p1)
```

#### Run Loop

```{r}

for(i in var_complete_ratio){
  list_b_ratio[[i]]<- update(mm,
                         formula=(paste0("Loc_bin ~", i, "+(1|Genus_species)+(1|gr(Taxon_Upham_style,cov = A))")),
                         family = bernoulli_missing,
                         newdata=dat,
                         refresh = 0,
                         cores = 4,
                         data2 = list(A = A),
                         prior = p1,
                         seed = 145
                         ) 
}

```

#### Save Models
```{r}
list_b_lm %>% write_rds(here("Data","B_lm_mods.Rds"))
list_b_ratio |> write_rds(here("Data","B_lm_mods.Rds"))
```

