---
title: "Binary Models — Individual Predictors with FL Rodents — Known Locomotion Scores"
output: html_notebook
---

# Overview

This script is a modified version of `Binary_Models.Rmd`. The key difference
is that this version uses a training dataset that includes known locomotion
scores (Loc_bin) for the four focal Florida rodent species:

- Ochrotomys nuttalli  → climbing (1)
- Peromyscus gossypinus → climbing (1)
- Peromyscus polionotus → no climbing (0)
- Podomys floridanus   → no climbing (0)

Because locomotion is now known for these species, the custom Bernoulli family
is not needed — standard bernoulli() is used instead, with Loc_bin as the
response directly. Everything else (model structure, priors, predictions,
plots) is identical to Binary_Models.Rmd.

---

# 1. Setup

```{r message = FALSE, warning=FALSE, include = FALSE}
pacman::p_load(tidyverse, brms, cmdstanr, kableExtra, here, ape, tidybayes, patchwork, glue)

options(brms.backend = "cmdstanr")

here::i_am("README.md")
here()
```

```{r}
# Load dataset with known locomotion scores for the FL species
dat <- read_csv(here("Data", "Clean_Data_FL_scores.csv"))

# Species in the dataset (needed to prune the phylogeny)
spp_obs <- dat$Taxon_Upham_style %>% unique()

# Variable lists — organized by measurement type and missingness
var_complete        <- read_rds(here("Data", "var_complete.Rds"))
var_missing         <- read_rds(here("Data", "var_missing.Rds"))
var_complete_linear <- read_rds(here("Data", "var_complete_linear.Rds"))
var_missing_linear  <- read_rds(here("Data", "var_missing_linear.Rds"))
var_complete_ratio  <- read_rds(here("Data", "var_complete_ratio.Rds"))
var_missing_ratio   <- read_rds(here("Data", "var_missing_ratio.Rds"))
```

```{r}
# Load phylogenetic tree and compute correlation matrix
tr <- ape::read.nexus(here("Data", "MCC_Upham_all_mammals.nex"))
tr <- ape::keep.tip(tr, spp_obs)
A  <- ape::vcv.phylo(tr, corr = TRUE)
```

---

# 2. Linear Measurement Models

Linear measurements are fit with log body mass as a covariate. Body mass is
itself sometimes missing, so it is also imputed via a submodel.

All predictors are z-scored prior to modeling (done in Data_Load.Rmd).

## 2a. Linear measurements with predictor missing

```{r}
list_b_lm <- vector(mode = "list")
```

```{r}
for (i in var_missing_linear) {
  
  # Standard bernoulli() — Loc_bin used directly as response
  bin_mod <- bf(
    paste0("Loc_bin ~ mi(", i, ") + mi(log_Mass) + 
            (1|Genus_species) + (1|gr(Taxon_Upham_style, cov = A))")
  ) + bernoulli()
  
  imp_mod <- bf(
    paste0(i, " | mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl +
            (1|gr(Taxon_Upham_style, cov = A))")
  ) + gaussian()
  
  mass_mod <- bf(
    "log_Mass | mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl +
     (1|gr(Taxon_Upham_style, cov = A))"
  ) + gaussian()
  
  m_form <- bin_mod + imp_mod + mass_mod + set_rescor(FALSE)
  
  # Priors set per response:
  # - Tighter SD prior for binary outcome (phylo/species signal is small)
  # - Wider SD prior for Gaussian imputation submodels
  pri <- get_prior(m_form, data2 = list(A = A), data = dat) %>%
    mutate(
      prior = case_when(
        class == "b" & coef == ""                      ~ "normal(0, 2)",
        class == "sd" & coef == "" & resp == "Locbin"  ~ "normal(0, 0.3)",
        class == "sd" & coef == "" & resp != "Locbin"  ~ "normal(0, 1)",
        class == "Intercept"                           ~ "normal(0, 2)",
        class == "sigma"                               ~ "normal(0, 0.5)",
        TRUE ~ prior
      )
    )
  
  list_b_lm[[i]] <- brm(
    m_form,
    data    = dat,
    data2   = list(A = A),
    prior   = pri,
    cores   = 4,
    seed    = 145,
    refresh = 0,
    silent  = 2
  )
}
```

## 2b. Linear measurements with complete data (only log_Mass missing)

```{r}
for (i in var_complete_linear) {
  
  bin_mod <- bf(
    paste0("Loc_bin ~ ", i, " + mi(log_Mass) + 
            (1|Genus_species) + (1|gr(Taxon_Upham_style, cov = A))")
  ) + bernoulli()
  
  mass_mod <- bf(
    "log_Mass | mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl +
     (1|gr(Taxon_Upham_style, cov = A))"
  ) + gaussian()
  
  m_form <- bin_mod + mass_mod + set_rescor(FALSE)
  
  pri <- get_prior(m_form, data2 = list(A = A), data = dat) %>%
    mutate(
      prior = case_when(
        class == "b" & coef == ""                      ~ "normal(0, 2)",
        class == "sd" & coef == "" & resp == "Locbin"  ~ "normal(0, 0.3)",
        class == "sd" & coef == "" & resp != "Locbin"  ~ "normal(0, 1)",
        class == "Intercept"                           ~ "normal(0, 2)",
        class == "sigma"                               ~ "normal(0, 0.5)",
        TRUE ~ prior
      )
    )
  
  list_b_lm[[i]] <- brm(
    m_form,
    data    = dat,
    data2   = list(A = A),
    prior   = pri,
    cores   = 4,
    seed    = 145,
    refresh = 0,
    silent  = 2
  )
}
```

---

# 3. Functional Ratio / Index Models

Ratios and indices are dimensionless so size is not included as a covariate.

## 3a. Ratios with missing data

```{r}
list_b_ratio <- vector(mode = "list")
```

```{r}
for (i in var_missing_ratio) {
  
  bin_mod <- bf(
    paste0("Loc_bin ~ mi(", i, ") + 
            (1|Genus_species) + (1|gr(Taxon_Upham_style, cov = A))")
  ) + bernoulli()
  
  imp_mod <- bf(
    paste0(i, " | mi() ~ Fl + Fdw + Fsw + Tpw + Hl + Hdw + Hpw + Hsw + Uol + Mtl +
            (1|gr(Taxon_Upham_style, cov = A))")
  ) + gaussian()
  
  m_form <- bin_mod + imp_mod + set_rescor(FALSE)
  
  pri <- get_prior(m_form, data2 = list(A = A), data = dat) %>%
    mutate(
      prior = case_when(
        class == "b" & coef == ""                      ~ "normal(0, 2)",
        class == "sd" & coef == "" & resp == "Locbin"  ~ "normal(0, 0.3)",
        class == "sd" & coef == "" & resp != "Locbin"  ~ "normal(0, 1)",
        class == "Intercept"                           ~ "normal(0, 2)",
        class == "sigma"                               ~ "normal(0, 0.5)",
        TRUE ~ prior
      )
    )
  
  list_b_ratio[[i]] <- brm(
    m_form,
    data    = dat,
    data2   = list(A = A),
    prior   = pri,
    cores   = 4,
    seed    = 145,
    refresh = 0,
    silent  = 2
  )
}
```

## 3b. Ratios with complete data

For complete ratio variables, no imputation submodel is needed. A single
initial fit is used with update() to avoid recompiling Stan for every variable.

```{r}
# Initial model to compile Stan — update() reuses the compiled code
p1 <- c(
  prior(normal(0, 2), class = Intercept),
  prior(normal(0, 2), class = b),
  prior(normal(0, 0.3), class = sd)
)

mm <- brm(
  Loc_bin ~ Fl + (1|Genus_species) + (1|gr(Taxon_Upham_style, cov = A)),
  data    = dat,
  data2   = list(A = A),
  family  = bernoulli(),
  prior   = p1,
  cores   = 4,
  refresh = 0
)
```

```{r}
for (i in var_complete_ratio) {
  list_b_ratio[[i]] <- update(
    mm,
    formula = paste0("Loc_bin ~ ", i,
                     " + (1|Genus_species) + (1|gr(Taxon_Upham_style, cov = A))"),
    newdata = dat,
    cores   = 4,
    seed    = 145,
    refresh = 0
  )
}
```

---

# 4. Save Models

```{r}
list_b_lm    %>% write_rds(here("Data", "B_lm_mods_FL_scores.Rds"))
list_b_ratio %>% write_rds(here("Data", "B_ratio_mods_FL_scores.Rds"))
```

---

# 5. Predictions

## 5a. Prepare prediction data

```{r}
mice <- c("Podomys_floridanus", "Ochrotomys_nuttalli",
          "Peromyscus_gossypinus", "Peromyscus_polionotus")

# These species have known Loc_bin so re_formula = NULL uses their
# fitted species random effects rather than drawing new levels
model_dat <- dat %>%
  filter(Genus_species %in% mice)
```

## 5b. Generate predictions from all models

```{r}
all_mods <- c(list_b_lm, list_b_ratio)

# For each model, generate predictions averaged over individuals
# resp = "Locbin" because brms strips underscores from "Loc_bin"
pred_all <- map_dfr(names(all_mods), function(var_name) {
  
  mod <- all_mods[[var_name]]
  
  if (is.null(mod)) {
    warning(glue("Model for {var_name} is NULL, skipping"))
    return(NULL)
  }
  
  tryCatch({
    mod %>%
      add_epred_draws(
        newdata    = model_dat,
        resp       = "Locbin",
        value      = "Prob_of_Climbing",
        re_formula = NULL
      ) %>%
      as.data.frame() %>%
      group_by(Genus_species, .draw) %>%
      summarise(
        Prob_of_Climbing = mean(Prob_of_Climbing, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(var = var_name)
  }, error = function(e) {
    warning(glue("Prediction failed for {var_name}: {e$message}"))
    return(NULL)
  })
})
```

## 5c. Plot function

```{r}
var_acc  <- c(names(list_b_lm), names(list_b_ratio))
n_shades <- floor(length(var_acc) / 2)

shades <- data.frame(
  xmin = seq(1.5, by = 2, length.out = n_shades),
  xmax = seq(2.5, by = 2, length.out = n_shades),
  ymin = -Inf,
  ymax = Inf
)

hplot <- function(groups) {
  pred_all %>%
    filter(Genus_species %in% groups) %>%
    mutate(
      var           = as.factor(var),
      Genus_species = gsub("_", " ", Genus_species)
    ) %>%
    ggplot(aes(
      y     = Prob_of_Climbing,
      x     = fct_relevel(var, var_acc),
      shape = Genus_species
    )) +
    geom_rect(
      inherit.aes = FALSE,
      data        = shades,
      mapping     = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      alpha       = 0.125
    ) +
    geom_hline(yintercept = 0.5, linetype = 2, linewidth = 0.35, color = "grey30") +
    stat_pointinterval(
      .width    = 0.8,
      color     = "grey15",
      linewidth = 1.7,
      position  = position_dodge(width = -0.8)
    ) +
    scale_y_continuous(limits = c(0, 1), expand = c(0.02, 0.02)) +
    scale_shape_manual(values = c(1, 15, 17, 19, 8)) +
    labs(x = "", y = "Probability of Climbing") +
    theme_pander(gM = FALSE) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.7),
      legend.text  = element_text(face = "italic"),
      legend.title = element_blank()
    ) +
    guides(shape = guide_legend(override.aes = list(size = 8)))
}
```

```{r}
hplot(mice)

ggsave(here("Plots", "Bin_Individual_Plots_FL_scores.pdf"), height = 8, width = 13)
```
