---
title: "Data Wrangle"
#output: html_notebook
---

# Data Wrangle

This notebook is for wrangling the data for subsequent analyses. This includes calculating functional indices, geometric means, log mass, etc. 

```{r}
#install.packages('tidyverse', 'here')
library(tidyverse)
library(here)
#this sets the working directory by telling it that the dir is the one containing "README.md"
#this way we can easily use different working directories between our machines.
here::i_am("README.md")
#this function will be used to scale variables to a mean of 0 and a standard deviation of 1
scale2 <- function(x, na.rm = TRUE) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
```

## Load and Wrangle Data

`read_csv` of the `readr` package reads in a dataframe as a tibble. 

I'll use your previous code to generate the indices, calculate geometric means, etc. But I'll do it all in one command, and by appending lines to the dataframe.
`mutate` creates a new varaible in the dataframe, but leaves the others untouched. 
```{r}
dat <-  read_csv(here("Data", "MouseMeasurements.csv")) |> 
  mutate(Genus_species = str_remove_all(Genus_species, " ")) |> 
  #this selects all columns from `catalog` to `Date`, removing the last ones
  select(catalog:Pppl) |> 
  # Make indices
  mutate(
    MANUS = Ppl/Mcl,      #MANUS Index
    HRI = Hsw/Hl,         #Humeral Robustness Index
    HPI = Hpw/Hl,         #Humeral Proximal Index
    HEB = Hdw/Hl,         #Humeral Epicondyle Breadth
    OLI = Uol/Ul,         #Olecranon Process Length Index
    BI = Rl/Hl,            #Brachial Index
    IM = (Hl+Ul)/(Fl+Tl), #Intermembral Index
    PRTI = Mcl/(Hl+Rl),   #Palm Robustness Index
    FRI = Fsw/Fl,         #Femoral Robustness
    FEB = Fdw/Fl,         #Femoral Epicondyle Breadth
    CI = Tl/Fl,           #Crural Index
    TRI = Tmw/Tl,         #Tibial Robustness Index
    PES = Pppl/Mtl,       #PES Index
    #Skipping geometric Mean
    #log mass
    log_Mass = log(mass_g)
        ) |> 
  #log transforms all linear variables
  mutate_at(vars(Fl:Pppl), log) |> 
  #creates a new column telling us this is the prediciton data
  mutate(pred_data = "T") |> 
  #removes this measurement
  select(!Usw)
```

## Load in Training data
This is the raw extant data for predictions
Some rows will need to be dropped and added. 

```{r}
tdat <- read_csv(here("Data", "Raw_Extant_Data.csv"), na = c("NA", "?", ""))|> 
   mutate(Loc_Ord = case_when(Loc_mode_Ordinal == "G"  ~ "G",
                             Loc_mode_Ordinal == "A"  ~ "A",
                             Loc_mode_Ordinal == "Sc" ~ "Sc",
                             Loc_mode_Ordinal == "T"  ~ "T",
                             Loc_mode_Ordinal == "Is" ~ "Is",
                             Loc_mode_Ordinal == "Sf" ~ "Sf",
                             Loc_mode_Ordinal == "Ss" ~ "Ss",
                             TRUE ~ NA),
         Loc_Ord = as.ordered(Loc_Ord),
         Loc_Ord = fct_relevel(Loc_Ord, c("G","A","Sc","T","Is","Sf","Ss")),
         # Combining the Intersurface and Semifossorial
         Loc_Ord2 = case_when(Loc_mode_Ordinal == "G" ~ "A",
                              Loc_mode_Ordinal == "A" ~ "A",
                              Loc_mode_Ordinal == "Sc" ~ "Sc",
                              Loc_mode_Ordinal == "T" ~ "T",
                              Loc_mode_Ordinal == "Is" ~ "B",
                              Loc_mode_Ordinal == "Sf" ~ "B",
                              Loc_mode_Ordinal == "Ss" ~ "B",
                             TRUE ~ NA),
         Loc_Ord2 = as.ordered(Loc_Ord2),
         Loc_Ord2 = fct_relevel(Loc_Ord2, c("B","T","Sc","A")),
         Loc_bin = case_when(Loc_mode_Binary == "Ground" ~ 0,
                             Loc_mode_Binary == "Tree" ~ 1,
                             TRUE ~ NA
                             ),
        # Loc_bin = as.factor(Loc_bin),
         Loc_mode_Categorical = as.factor(Loc_mode_Categorical)) |> 
#################
#Calculate Indices!
#################
# takes the measurements, and turns them into indices. See preprint for more info.
   mutate(
    MANUS = Ppl/Mcl,      #MANUS Index
    HRI = Hsw/Hl,         #Humeral Robustness Index
    HPI = Hpw/Hl,         #Humeral Proximal Index
    HEB = Hdw/Hl,         #Humeral Epicondyle Breadth
    OLI = Uol/Ul,         #Olecranon Process Length Index
    BI = Rl/Hl,            #Brachial Index
    IM = (Hl+Ul)/(Fl+Tl), #Intermembral Index
    PRTI = Mcl/(Hl+Rl),   #Palm Robustness Index
    FRI = Fsw/Fl,         #Femoral Robustness
    FEB = Fdw/Fl,         #Femoral Epicondyle Breadth
    CI = Tl/Fl,           #Crural Index
    TRI = Tmw/Tl,         #Tibial Robustness Index
    PES = Pppl/Mtl,       #PES Index
    #log mass
    log_Mass = log(Mass_grams)
        ) |> 
  #log transforms all linear variables
  mutate_at(vars(Skl:Pdpw), log) |> 
  relocate(Loc_bin, .after = Loc_mode_Binary) |> 
  relocate(Loc_Ord, .after = Loc_mode_Ordinal) |> 
  relocate(Loc_Ord2, .after = Loc_Ord) |>
  relocate(log_Mass, .before = Skl) |> 
  rename(catalog = Specimen_Num) |> 
  mutate(catalog = str_remove_all(catalog, " ")) |> 
  mutate(pred_data = "F") |> 
  select(Genus_species:log_Mass, Hl:Pdpw, MANUS:pred_data)
```


```{r}
setdiff(names(dat), names(tdat))
setdiff(names(tdat), names(dat))
```
## Combine Datasets

Scale them to a mean of 0 and a standard deviation of 1. 
Then save it. This will be the main dataset!


```{r}
dat2 <- bind_rows(dat, tdat) |> 
  #Removing all the measurements in the big dataset that don't matter here
  select(!c("Hhw", "Hdcw", "Hhl", "Pel", "Il", "Isl", "Fhd", "Fgh", "Tdw", "Fbl", "Fbmw", "Fbpw", "Fbdw", "Csw", "Ccw", "Cl", "Ctl", "Ctw", "Cal", "Anl", "Al", "Atw", "Mcw", "Ppw", "Ipl", "Ipw", "Dpl", "Dpw", "Mtw", "Pppw", "Pipl", "Pipw", "Pdpl", "Pdpw")) |>
   mutate_at(vars(Fl:log_Mass), scale2) |> 
  write_csv(here("Data", "Clean_Data.csv"))
```

## Other Pieces of the Puzzle

What data are missing?

```{r}
dat2 %>% summarise(across(Fl:log_Mass, ~sum(is.na(.x)))) %>% 
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>% 
  arrange(n_NA) |> print()

# what from the FL dataset is missing?

dat2 %>% filter(pred_data == "T") |> 
  summarise(across(Fl:log_Mass, ~sum(is.na(.x)))) %>% 
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>% 
  arrange(n_NA) |> 
  filter(n_NA == 1) |> print()
```

The complete data are only missing 1 specimen from the new set. 
The var_missing are missing a few, and need to be estimated
Problem is that we are missing mass from quite a few, so we need to estimate. All models will be missing data models, except "HRI", "HPI", "HEB", "FRI", and "FEB"
```{r}
var_complete <- c("Fl", "Fdw", "Fsw", "Tpw", "Hl", "Hdw", "Hpw", "Hsw", "HRI", "HPI", "HEB", "FRI", "FEB", "Uol", "Mtl")|> write_rds(here("Data", "var_complete.Rds"))

var_missing <- c("Tmw", "Mcl", "Rl", "OLI", "BI", "Ppl", "Pppl",  "Ul", "PES", "MANUS", "PRTI", "log_Mass", "Tl", "CI", "TRI", "IM") |> write_rds(here("Data", "var_missing.Rds"))

var_complete_linear <- c("Fl", "Fdw", "Fsw", "Tpw", "Hl", "Hdw", "Hpw", "Hsw", "Uol", "Mtl")|> 
  write_rds(here("Data", "var_complete_linear.Rds"))

# Removing log_mass here beacuse we will have to estimate it always
var_missing_linear <- c("Tmw", "Mcl", "Ul", "Rl",  "Ppl", "Pppl",  "Tl") |>
  write_rds(here("Data", "var_missing_linear.Rds"))

var_complete_ratio <- c("HRI", "HPI", "HEB", "FRI", "FEB") |> 
   write_rds(here("Data", "var_complete_ratio.Rds"))

var_missing_ratio <- c( "OLI", "BI", "PES", "MANUS", "PRTI", "CI", "TRI", "IM") |> 
  write_rds(here("Data", "var_missing_ratio.Rds"))

var_accurate <- c('HPI', 'PES', 'MANUS', 'Ppl', 'Hl', 'OLI', 'Rl', 'Uol', 'Fl', 'Ul') |> 
  write_rds(here("Data", "var_accurate.Rds"))
```


## Preliminary Plots
Using the data to make some prelim plots
Adding the mean of each species to the plots

MANUS
```{r}
dat |> 
  #lets visualize the means of each species
  group_by(Genus_species) |> 
  mutate(mean_MANUS = mean(MANUS, na.rm = TRUE)) |> 
  #now begin plotting
  ggplot(aes(x = MANUS, y = Genus_species)) + 
  geom_point() +
  geom_point(aes(x = mean_MANUS), 
             color = "red", 
             size = 5, 
             #transparency
             alpha = 0.1) +
  labs(title = "MANUS") +
  theme_bw()
```
HPI
```{r}
dat |> 
  #lets visualize the means of each species
  group_by(Genus_species) |> 
  mutate(mean_HPI = mean(HPI, na.rm = TRUE)) |> 
  #now begin plotting
  ggplot(aes(x = HPI, y = Genus_species)) + 
  geom_point() +
  geom_point(aes(x = mean_HPI), 
             color = "red", 
             size = 5, 
             #transparency
             alpha = 0.1) +
  labs(title = "HPI") +
  theme_bw()
```

OLI

```{r}
dat |> 
  #lets visualize the means of each species
  group_by(Genus_species) |> 
  mutate(mean_OLI = mean(OLI, na.rm = TRUE)) |> 
  #now begin plotting
  ggplot(aes(x = OLI, y = Genus_species)) + 
  geom_point() +
  geom_point(aes(x = mean_OLI), 
             color = "red", 
             size = 5, 
             #transparency
             alpha = 0.1) +
  labs(title = "OLI") +
  theme_bw()
```

PES
```{r}
dat |> 
  #lets visualize the means of each species
  group_by(Genus_species) |> 
  mutate(mean_PES = mean(PES, na.rm = TRUE)) |> 
  #now begin plotting
  ggplot(aes(x = PES, y = Genus_species)) + 
  geom_point() +
  geom_point(aes(x = mean_PES), 
             color = "red", 
             size = 5, 
             #transparency
             alpha = 0.1) +
  labs(title = "PES") +
  theme_bw()
```

