---
title: "Morphological Distributions by Species"

---

# Overview

This script fits a series of Bayesian distributional regression models to
estimate the distribution of each skeletal measurement across four focal
Florida rodent species. One model is fit per variable in `var_accurate`.

All variables use the same model structure:
  - mu    ~ 0 + Genus_species
  - sigma ~ 0 + Genus_species

Size-standardized linear measurements are represented as log-shape ratios
(e.g. lsPpl = log(Ppl / gm)), which remove the effect of body size by
construction. This unifies the model framework — no separate mass imputation
or size-correction step needed.

---

# 1. Setup

```{r message = FALSE, warning=FALSE, include = FALSE}
pacman::p_load(tidyverse, brms, cmdstanr, here, tidybayes, 
               patchwork, glue, ggthemes)

options(brms.backend = "cmdstanr")

here::i_am("README.md")
here()
```

```{r}
dat <- read_csv(here("Data", "MouseMeasurements.csv")) |> 
  mutate(Genus_species = str_remove_all(Genus_species, " ")) |> 
  select(catalog:Pppl) |> 
  mutate(
    MANUS = Ppl/Mcl,
    HRI   = Hsw/Hl,
    HPI   = Hpw/Hl,
    HEB   = Hdw/Hl,
    OLI   = Uol/Ul,
    BI    = Rl/Hl,
    IM    = (Hl+Ul)/(Fl+Tl),
    PRTI  = Mcl/(Hl+Rl),
    FRI   = Fsw/Fl,
    FEB   = Fdw/Fl,
    CI    = Tl/Fl,
    TRI   = Tmw/Tl,
    PES   = Pppl/Mtl
  ) |> 
  mutate(gm = (Hl * Fl * Ul * Rl) ^ (1/4),
         HBL = TotL - Tail) |> 
  # Log-shape ratios: size-standardized linear measurements
  mutate(
    lsHl   = log(Hl   / gm),
    lsHsw  = log(Hsw  / gm),
    lsHdw  = log(Hdw  / gm),
    lsHpw  = log(Hpw  / gm),
    lsUol  = log(Uol  / gm),
    lsUl   = log(Ul   / gm),
    lsRl   = log(Rl   / gm),
    lsFl   = log(Fl   / gm),
    lsFsw  = log(Fsw  / gm),
    lsFdw  = log(Fdw  / gm),
    lsTmw  = log(Tl   / gm),
    lsPppl = log(Pppl / gm),
    lsMtl  = log(Mtl  / gm),
    lsPpl  = log(Ppl  / gm)
  ) |> 
  #mutate_at(vars(Fl:Pppl), log) |>
  mutate(pred_data = "T") |> 
  select(!Usw) |>
  filter(catalog != "U20988*") |> 
  mutate(Genus_species = factor(Genus_species, levels = mice))
```

```{r}
# Variable sets
var_accurate <- read_rds(here("Data", "var_accurate.Rds"))

# Log-shape ratio versions of formerly size-controlled variables
vars_lsr <- c("lsPpl", "lsHl", "lsRl", "lsUol", "lsFl", "lsUl")

# Dimensionless indices — unchanged
vars_idx <- c("HPI", "PES", "MANUS", "OLI")

# Combined variable set for modeling
vars_all <- c(vars_lsr, vars_idx)

# Restrict to those present in var_accurate if needed
# (comment out if var_accurate doesn't include ls-prefixed names)
# vars_all <- vars_all[vars_all %in% var_accurate]

# Color palette
pal <- c('#62361B', '#f48a1a', '#e6ab02', '#66a61e', '#31c1bb', 
         '#0d57ff', '#5e1ea6', '#f9581c', '#f384f8')
```

---

# 2. Fit Models

Single distributional regression framework for all variables.
`mu` and `sigma` both vary by species. Using `0 + Genus_species` suppresses
the intercept so each species gets its own coefficient directly — no
reference-level contrast. This makes posterior extraction straightforward.

```{r}
p <- c(
  prior(normal(0, 2),   class = b, dpar = ""),
  prior(normal(0, 0.5), class = b, dpar = "sigma")
)

dist_mods <- vector(mode = "list")

# Compile initial model — update() reuses this Stan executable for all variables
mm <- brm(
  bf(HPI ~ 0 + Genus_species, sigma ~ 0 + Genus_species),
  data    = dat,
  family  = gaussian(),
  prior   = p,
  cores   = 4,
  seed    = 145,
  refresh = 0,
  silent  = 2
)

for (i in vars_all) {
  
  n_obs <- dat %>% pull(all_of(i)) %>% na.omit() %>% length()
  if (n_obs < 10) {
    warning(glue("Skipping {i}: only {n_obs} non-NA observations"))
    next
  }
  
  dist_mods[[i]] <- update(
    mm,
    formula = bf(
      as.formula(paste0(i, " ~ 0 + Genus_species")),
      as.formula("sigma ~ 0 + Genus_species")
    ),
    newdata = dat,
    cores   = 4,
    seed    = 145,
    refresh = 0,
    silent  = 2
  )
}
```

```{r}
dist_mods %>% write_rds(here("Data", "dist_mods.Rds"))
```

---

# 3. Extract Posterior Draws

With `0 + Genus_species` parameterization, each species has its own named
coefficient. We extract these directly using `spread_draws()` — no prediction
step needed.

```{r}
post_all <- map_dfr(names(dist_mods), function(var_name) {
  
  mod <- dist_mods[[var_name]]
  if (is.null(mod)) return(NULL)
  
  newdata <- dat %>% 
    distinct(Genus_species) %>% 
    filter(!is.na(Genus_species))
  
  tryCatch({
    mod %>%
      predicted_draws(newdata = newdata) %>%
      as.data.frame() %>%
      rename(value = .prediction) %>%
      mutate(var = var_name) %>%
      select(.draw, Genus_species, value, var)
  }, error = function(e) {
    warning(glue("Draw extraction failed for {var_name}: {e$message}"))
    return(NULL)
  })
})

```


# 3b. Measurement Score Card

```{r}
post_all %>%
  group_by(var, Genus_species) %>%
  summarise(mean = mean(value), .groups = "drop") %>%
  pivot_wider(names_from = Genus_species, values_from = mean) %>%
  mutate(Ochrotomys_diff = Ochrotomys_nuttalli - rowMeans(across(c(Podomys_floridanus, 
                                                                    Peromyscus_gossypinus, 
                                                                    Peromyscus_polionotus))))
```

---

# 4. Plot

```{r}
# Variable order for plotting
# If var_accurate has ls-prefixed names, use it; otherwise use vars_all order
var_plot_order <- vars_all  # adjust if var_accurate contains the ls names

n_shades <- floor(length(var_plot_order) / 2)

shades <- data.frame(
  xmin = seq(1.5, by = 2, length.out = n_shades),
  xmax = seq(2.5, by = 2, length.out = n_shades),
  ymin = -Inf,
  ymax = Inf
)

# Raw data in long format for jittered points
dat_long <- dat %>%
  select(Genus_species, catalog, all_of(var_plot_order)) %>%
  pivot_longer(
    cols      = all_of(var_plot_order),
    names_to  = "var",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  group_by(var) %>%
  mutate(value = scale(value)[, 1]) %>%
  ungroup() %>%
  mutate(Genus_species = gsub("_", " ", as.character(Genus_species)))
```



```{r}
morpho_plot <- function(groups) {
  
  ppred_plot <- post_all %>%
    filter(Genus_species %in% groups) %>%
    mutate(Genus_species = gsub("_", " ", as.character(Genus_species)))
  
  raw_plot <- dat_long %>%
    filter(gsub(" ", "_", Genus_species) %in% groups)
  
  ppred_plot %>%
    ggplot(aes(
      x     = fct_relevel(var, var_plot_order),
      y     = value,
      shape = Genus_species
    )) +
    geom_rect(
      inherit.aes = FALSE,
      data        = shades,
      mapping     = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      alpha       = 0.125
    ) +
    stat_pointinterval(
      .width    = 0.8,
      color     = "grey15",
      linewidth = 1.7,
      position  = position_dodge(width = 0.5)
    ) +
    geom_point(
      data    = raw_plot,
      mapping = aes(x = fct_relevel(var, var_plot_order), y = value,
                    shape = Genus_species),
      position = position_jitterdodge(dodge.width = 0.5, jitter.width = 0.1,
                                      jitter.height = 0),
      size  = 1.5,
      alpha = 0.5,
      color = "grey30"
    ) +
    scale_shape_manual(values = c(1, 15, 17, 19, 8)) +
    labs(x = "", y = "Z-score") +
    theme_pander(gM = FALSE) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.7),
      legend.text  = element_text(face = "italic"),
      legend.title = element_blank()
    ) +
    guides(shape = guide_legend(override.aes = list(size = 4)))
}

morpho_plot_clean <- function(groups) {
  
  ppred_plot <- post_all %>%
    filter(Genus_species %in% groups) %>%
    mutate(Genus_species = gsub("_", " ", as.character(Genus_species)))
  
  ppred_plot %>%
    ggplot(aes(
      x     = fct_relevel(var, var_plot_order),
      y     = value,
      shape = Genus_species
    )) +
    geom_rect(
      inherit.aes = FALSE,
      data        = shades,
      mapping     = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      alpha       = 0.125
    ) +
    stat_pointinterval(
      .width    = 0.8,
      color     = "grey15",
      linewidth = 1.7,
      position  = position_dodge(width = 0.5)
    ) +
    scale_shape_manual(values = c(1, 15, 17, 19, 8)) +
    labs(x = "", y = "Z-score") +
    theme_pander(gM = FALSE) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.7),
      legend.text  = element_text(face = "italic"),
      legend.title = element_blank()
    ) +
    guides(shape = guide_legend(override.aes = list(size = 4)))
}
```

```{r}
morpho_plot(mice)
ggsave(here("Plots", "Morpho_Distributions.pdf"), height = 8, width = 13)

morpho_plot_clean(mice)
ggsave(here("Plots", "Morpho_Distributions_no_points.pdf"), height = 8, width = 13)
```


```{r}
# Faceted by variable — 2x5 grid, one panel per variable, species on x-axis
morpho_plot_byvar <- function(groups) {
  
  ppred_plot <- post_all %>%
    filter(Genus_species %in% groups) %>%
    mutate(Genus_species = gsub("_", " ", as.character(Genus_species)))
  
  raw_plot <- dat_long %>%
    filter(gsub(" ", "_", Genus_species) %in% groups)
  
  ppred_plot %>%
    ggplot(aes(
      x     = Genus_species,
      y     = value,
      shape = Genus_species
    )) +
    stat_pointinterval(
      .width    = 0.8,
      color     = "grey15",
      linewidth = 1.7,
      position  = position_dodge(width = 0.5)
    ) +
    scale_shape_manual(values = c(1, 15, 17, 19, 8)) +
    facet_wrap(~ var, scales = "free_y", nrow = 2) +
    labs(x = "", y = "Z-score") +
    theme_pander(gM = FALSE) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1, face = "italic", size = 7),
      panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.7),
      legend.position = "none",
      strip.background = element_blank()
    )
}

morpho_plot_byvar(mice)
ggsave(here("Plots", "Morpho_Distributions_byvar.pdf"), height = 8, width = 13)
```

#5. Post Means and Contrasts

```{r}
post_means <- map_dfr(names(dist_mods), function(var_name) {
  dist_mods[[var_name]] %>%
    spread_draws(`b_Genus_species.*`, regex = TRUE) %>%
    pivot_longer(
      starts_with("b_Genus_species"),
      names_to  = "Genus_species",
      values_to = "value"
    ) %>%
    mutate(
      Genus_species = str_remove(Genus_species, "b_Genus_species"),
      var           = var_name
    ) %>%
    select(.draw, Genus_species, value, var)
})

contrasts <- post_means %>%
  pivot_wider(names_from = Genus_species, values_from = value) %>%
  group_by(var) %>%
  summarise(
    vs_Podomys    = mean(Ochrotomys_nuttalli > Podomys_floridanus),
    vs_gossypinus = mean(Ochrotomys_nuttalli > Peromyscus_gossypinus),
    vs_polionotus = mean(Ochrotomys_nuttalli > Peromyscus_polionotus)
  ) %>%
  mutate(
    predicted_higher = var %in% c("MANUS", "PES", "lsHl", "lsPpl", "lsRl", "lsUl"),
    across(starts_with("vs_"), ~ if_else(predicted_higher, .x, 1 - .x),
           .names = "{.col}_directional")
  )
```


```{r}
pairs <- combn(unique(post_means$Genus_species), 2, simplify = FALSE)

diffs_all <- map_dfr(pairs, function(p) {
  post_means %>%
    filter(Genus_species %in% p) %>%
    pivot_wider(names_from = Genus_species, values_from = value) %>%
    mutate(diff = .data[[p[1]]] - .data[[p[2]]],
           comparison = paste(p[1], "vs", p[2])) %>%
    select(var, .draw, diff, comparison)
}) %>%
  mutate(comparison = str_replace_all(comparison, 
                                      "Peromyscus_|Ochrotomys_|Podomys_", ""))

diffs_all %>%
  ggplot(aes(x = diff, y = comparison)) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
  stat_pointinterval(.width = 0.89) +
  facet_wrap(~ var, nrow = 2, scales = "free_x") +
  labs(x = "Posterior difference", y = "") +
  theme_bw() +
  theme(strip.background = element_blank(),
        panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.7))
```

