---
title: "Morphological Distributions by Species"
output: html_notebook
---

# Overview

This script fits a series of Bayesian distributional regression models to
estimate the distribution of each skeletal measurement across four focal
Florida rodent species. One model is fit per variable in `var_accurate`.

Model structure varies by variable type:

Size-controlled variables (Ppl, Hl, Rl, Uol, Fl, Ul):
  - mu    ~ 0 + Genus_species + log_Mass
  - sigma ~ 0 + Genus_species
  - Missing log_Mass values are filled in beforehand using a standalone
    mass prediction model fit to the full dataset

Non-size-controlled variables (HPI, PES, MANUS, OLI):
  - mu    ~ Genus_species
  - sigma ~ Genus_species

Predictions use species mean log_Mass for the size-controlled variables,
so intervals reflect morphological variation at a typical body size for
each species rather than being confounded by mass differences.

---

# 1. Setup

```{r message = FALSE, warning=FALSE, include = FALSE}
pacman::p_load(tidyverse, brms, cmdstanr, here, tidybayes, 
               patchwork, glue, ggthemes)

options(brms.backend = "cmdstanr")

here::i_am("README.md")
here()
```

```{r}
dat <-  read_csv(here("Data", "MouseMeasurements.csv")) |> 
  mutate(Genus_species = str_remove_all(Genus_species, " ")) |> 
  #this selects all columns from `catalog` to `Date`, removing the last ones
  select(catalog:Pppl) |> 
  # Make indices
  mutate(
    MANUS = Ppl/Mcl,      #MANUS Index
    HRI = Hsw/Hl,         #Humeral Robustness Index
    HPI = Hpw/Hl,         #Humeral Proximal Index
    HEB = Hdw/Hl,         #Humeral Epicondyle Breadth
    OLI = Uol/Ul,         #Olecranon Process Length Index
    BI = Rl/Hl,            #Brachial Index
    IM = (Hl+Ul)/(Fl+Tl), #Intermembral Index
    PRTI = Mcl/(Hl+Rl),   #Palm Robustness Index
    FRI = Fsw/Fl,         #Femoral Robustness
    FEB = Fdw/Fl,         #Femoral Epicondyle Breadth
    CI = Tl/Fl,           #Crural Index
    TRI = Tmw/Tl,         #Tibial Robustness Index
    PES = Pppl/Mtl) |>        #PES Index
  mutate(gm = ( Fl * Fdw * Fsw * Tpw * Hl * Hdw * Hpw * Hsw) ^ (1/8)) |> 
  #log transforms all linear variables
  mutate_at(vars(Fl:Pppl), log) |>
  #creates a new column telling us this is the prediciton data
  mutate(pred_data = "T") |> 
  #removes this measurement
  select(!Usw) %>%
  filter(catalog != "U20988*") |> 
  mutate(Genus_species = factor(Genus_species, levels = mice))
   
```


```{r}
# Variable sets
var_accurate <- read_rds(here("Data", "var_accurate.Rds"))

# Split into size-controlled and non-size-controlled
# Size-controlled: linear measurements where body size is a meaningful covariate
# Non-size-controlled: ratios/indices that are dimensionless by construction
vars_size_controlled <- c("Ppl", "Hl", "Rl", "Uol", "Fl", "Ul")
vars_no_size         <- c("HPI", "PES", "MANUS", "OLI")

# Keep only those present in var_accurate
vars_size_controlled <- vars_size_controlled[vars_size_controlled %in% var_accurate]
vars_no_size         <- vars_no_size[vars_no_size %in% var_accurate]

#"Fl", "Fdw", "Tpw", "Hl", "Hdw", "Hpw", "Hsw", "log_Mass"
# Color palette
pal <- c('#62361B', '#f48a1a', '#e6ab02', '#66a61e', '#31c1bb', 
         '#0d57ff', '#5e1ea6', '#f9581c', '#f384f8')
```

---

# 2. Fit Models

## 2a.

Standalone mass prediction model

Before fitting the size-controlled morphology models, we need to fill in
missing mass values for focal specimens. A standalone regression model
is fit to the full dataset (all species) to predict log_Mass from skeletal
measurements. Posterior means are then used to impute missing values in
dat_mice — simpler than the full mi() joint model and sufficient here since
we only need a point estimate per specimen.

```{r}
# Fit mass model to full dataset so all species inform the regression
# Predictors are the same complete skeletal measurements used elsewhere

mass_mod <- brm(
  mass_g ~ Hl + Hsw + Hdw + Uol + Ul + Rl + Fl + Fsw + Fdw,
  family  = gaussian(),
  data    = dat,       # full dataset, not just focal species
 # prior   = p_mass,
  cores   = 4,
  seed    = 145,
  refresh = 0,
  silent  = 2
)

```

```{r}
# Identify focal specimens with missing log_Mass
missing_mass <- dat %>%
  filter(is.na(mass_g)) %>%
  select(catalog, Genus_species, Hl, Hsw, Hdw, Uol, Ul, Rl, Fl, Fsw, Fdw)

# Get posterior mean log_Mass predictions for those specimens
# add_epred_draws() gives the expected value (mu), averaged over draws
mass_imputed <- mass_mod %>%
  add_epred_draws(newdata = missing_mass) %>%
  as.data.frame() %>%
  group_by(catalog) %>%
  summarise(mass_g_imputed = mean(.epred), .groups = "drop")

# Fill in missing log_Mass values with imputed posterior means
dat_mice <- dat %>%
  left_join(mass_imputed, by = "catalog") %>%
  mutate(mass_g = coalesce(mass_g, mass_g_imputed)) %>%
  select(-mass_g_imputed) |>
  mutate(log_Mass = log(mass_g), 
         log_gm = log(gm),
         lsHl = log(Hl/gm),
         lsHsw = log(Hsw/gm),
         lsHdw = log(Hdw/gm),
         lsHpw = log(Hpw/gm),
         lsUol = log(Uol/gm),
         lsUl = log(Ul/gm),
         lsRl = log(Rl/gm),
         lsFl = log(Fl/gm),
         lsFsw = log(Fsw/gm),
         lsFdw = log(Fdw/gm),
         lsTmw = log(Tl/gm),
         lsPppl = log(Pppl/gm),
         lsMtl = log(Mtl/gm),
         lsPpl = log(Ppl/gm)) |> 
    #log transforms all linear variables
  mutate_at(vars(Fl:Pppl), log) 

dat_mice |> ggplot(aes(x = Genus_species, y = lsPpl)) + 
  geom_point()
```

Simple distributional regression: mu and sigma both vary by species.
Using 0 + Genus_species suppresses the intercept so each species gets its
own coefficient directly — no reference-level contrast. This makes posterior
extraction straightforward: one b_Genus_species* column per species.

```{r}
# With 0 + parameterization there is no Intercept class — all species
# means are in class = b. sigma still gets its own Intercept (for the
# reference species) plus b coefficients for the others.
p_no_size <- c(
  prior(normal(0, 2),   class = b, dpar = ""),
  prior(normal(0, 0.5), class = b, dpar = "sigma")
)

dist_mods_no_size <- vector(mode = "list")

# Compile initial model — update() reuses this for all non-size variables
mm_no_size <- brm(
  bf(HPI ~ 0 + Genus_species, sigma ~ 0 + Genus_species),
  data    = dat_mice,
  family  = gaussian(),
  prior   = p_no_size,
  cores   = 4,
  seed    = 145,
  refresh = 0,
  silent  = 2
)

for (i in vars_no_size) {
  
  n_obs <- dat_mice %>% pull(all_of(i)) %>% na.omit() %>% length()
  if (n_obs < 10) {
    warning(glue("Skipping {i}: only {n_obs} non-NA observations"))
    next
  }
  
  dist_mods_no_size[[i]] <- update(
    mm_no_size,
    formula = bf(
      as.formula(paste0(i, " ~ 0 + Genus_species")),
      as.formula("sigma ~ 0 + Genus_species")
    ),
    newdata = dat_mice,
    cores   = 4,
    seed    = 145,
    refresh = 0,
    silent  = 2
  )
}
```

## 2b. 
Simple distributional regression: mu and sigma both vary by species.
Using 0 + Genus_species suppresses the intercept so each species gets its
own coefficient directly — no reference-level contrast. This makes posterior
extraction straightforward: one b_Genus_species* column per species.


```{r}

p <- c(
  prior(normal(0, 1), class = b, dpar = ""),
  prior(normal(0, 0.5), class = b, dpar = "sigma"))

m0 <- brm(bf(log_Mass ~  0 + Genus_species, sigma ~ 0 + Genus_species),
          data = dat_mice, 
          prior = p,
          cores   = 4,
          seed    = 145,
          refresh = 0,
          silent  = 2,
          refresh = 0)

m1 <-  update(m0, bf(HPI ~  0 + Genus_species, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m2 <- update(m0, bf(PES ~  0 + Genus_species, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m3 <- update(m0, bf(MANUS ~  0 + Genus_species, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m4 <- update(m0, bf(OLI ~  0 + Genus_species, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m5 <- update(m0, bf(Ppl ~  0 + Genus_species + log_Mass, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m6 <- update(m0, bf(Hl ~  0 + Genus_species + log_Mass, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m7 <- update(m0, bf(Rl ~  0 + Genus_species + log_Mass, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m8 <- update(m0, bf(Uol ~  0 + Genus_species + log_Mass, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m9 <- update(m0, bf(Fl ~  0 + Genus_species + log_Mass, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

m10 <- update(m0, bf(Ul ~  0 + Genus_species + log_Mass, sigma ~ 0 + Genus_species),
              newdata = dat_mice, prior = p, cores= 4, seed = 145, refresh = 0, silent  = 2)

```

```{r}
d <- bind_rows("log_Mass" = posterior_samples(m0),
               "HPI" = posterior_samples(m1), 
               "PES" = posterior_samples(m2),
               "MANUS" = posterior_samples(m3),
               "OLI" = posterior_samples(m4),
               "Ppl" = posterior_samples(m5),
               "Hl" = posterior_samples(m6),
               "Rl" = posterior_samples(m7),
               "Uol" = posterior_samples(m8),
               "Fl" = posterior_samples(m9),
               "Ul" = posterior_samples(m10),
               .id = "var") %>% 
  select(!c(lprior, lp__, b_log_Mass, starts_with("b_sigma"))) %>% 
  rename_with(str_sub, start = 16L, .cols = starts_with("b_Genus_species")) %>% 
  #mutate_at(15:26, scale2, na.rm = TRUE) %>% 
  pivot_longer(!var, names_to = "species", values_to = "val") %>% 
  group_by(var) %>% 
  mutate(val = scale2(val)) %>% ungroup() 
```

```{r}
d %>% 
    ggplot(aes(x = species, y = val)) +
    stat_pointinterval(.width = c(.60, .89), 
                       point_size = 1.5) +
    geom_hline(yintercept = 0, linetype = 3, size = 0.3) +
    scale_y_continuous(labels = scales::label_number(accuracy = 0.1))+
    facet_wrap( ~var,
               scales = 'free', ncol = 4) + 
    theme_bw() +
    labs(y = "Measurement Estimate (Scaled)", x = "") +
    theme(axis.text.x= element_text(angle=35,hjust=1, face = "italic"),
          strip.background = element_blank()) 
```


```{r}
# With 0 + parameterization there is no Intercept class — all species
# means are in class = b. sigma still gets its own Intercept (for the
# reference species) plus b coefficients for the others.
p_no_size <- c(
  prior(normal(0, 2),   class = b, dpar = ""),
  prior(normal(0, 0.5), class = b, dpar = "sigma")
)

dist_mods_no_size <- vector(mode = "list")

# Compile initial model — update() reuses this for all non-size variables
mm_no_size <- brm(
  bf(HPI ~ 0 + Genus_species, sigma ~ 0 + Genus_species),
  data    = dat_mice,
  family  = gaussian(),
  prior   = p_no_size,
  cores   = 4,
  seed    = 145,
  refresh = 0,
  silent  = 2
)

for (i in vars_no_size) {
  
  n_obs <- dat_mice %>% pull(all_of(i)) %>% na.omit() %>% length()
  if (n_obs < 10) {
    warning(glue("Skipping {i}: only {n_obs} non-NA observations"))
    next
  }
  
  dist_mods_no_size[[i]] <- update(
    mm_no_size,
    formula = bf(
      as.formula(paste0(i, " ~ 0 + Genus_species")),
      as.formula("sigma ~ 0 + Genus_species")
    ),
    newdata = dat_mice,
    cores   = 4,
    seed    = 145,
    refresh = 0,
    silent  = 2
  )
}

```


## 2c. Size-controlled distributional models

Now that log_Mass is complete for all focal specimens, these models are
straightforward — no mi() needed. Using 0 + Genus_species so each species
gets its own intercept directly, with log_Mass as a shared slope covariate.
update() works since all share the same formula structure.

```{r}
p_size <- c(
  prior(normal(0, 2),   class = b,         dpar = ""),
  prior(normal(0, 0.5), class = b,         dpar = "sigma")
)

dist_mods_size <- vector(mode = "list")

# Compile initial model — update() reuses this for all size-controlled variables
mm_size <- brm(
  bf(Fl ~ 0 + Genus_species + log_Mass, sigma ~ 0 + Genus_species),
  data    = dat_mice,
  family  = gaussian(),
  prior   = p_size,
  cores   = 4,
  seed    = 145,
  refresh = 0,
  silent  = 2
)

for (i in vars_size_controlled) {
  
  n_obs <- dat_mice %>% pull(all_of(i)) %>% na.omit() %>% length()
  if (n_obs < 10) {
    warning(glue("Skipping {i}: only {n_obs} non-NA observations"))
    next
  }
  
  dist_mods_size[[i]] <- update(
    mm_size,
    formula = bf(
      as.formula(paste0(i, " ~ 0 + Genus_species + log_Mass")),
      as.formula("sigma ~ 0 + Genus_species")
    ),
    newdata = dat_mice,
    cores   = 4,
    seed    = 145,
    refresh = 0,
    silent  = 2
  )
}
```

```{r}
mass_mod          %>% write_rds(here("Data", "mass_mod.Rds"))
dist_mods_no_size %>% write_rds(here("Data", "dist_mods_no_size.Rds"))
dist_mods_size    %>% write_rds(here("Data", "dist_mods_size.Rds"))
```

---

# 3. Extract Posterior Draws

With 0 + Genus_species parameterization, each species has its own named
coefficient (b_Genus_speciesOchrotomys_nuttalli, etc.). We extract these
directly using spread_draws() — no prediction step needed. This gives the
posterior distribution of the species mean for each variable, already
size-adjusted for the size-controlled variables.

The species name cleanup (removing "b_Genus_species" prefix) is done with
str_remove() to recover clean Genus_species labels for plotting.

## 3a. Non-size-controlled variables

```{r}
post_no_size <- map_dfr(names(dist_mods_no_size), function(var_name) {
  
  mod <- dist_mods_no_size[[var_name]]
  if (is.null(mod)) return(NULL)
  
  tryCatch({
    mod %>%
      spread_draws(`b_Genus_species.*`, regex = TRUE) %>%
      pivot_longer(
        starts_with("b_Genus_species"),
        names_to  = "Genus_species",
        values_to = "value"
      ) %>%
      mutate(
        Genus_species = str_remove(Genus_species, "b_Genus_species"),
        var           = var_name
      ) %>%
      select(.draw, Genus_species, value, var)
  }, error = function(e) {
    warning(glue("Draw extraction failed for {var_name}: {e$message}"))
    return(NULL)
  })
})
```

## 3b. Size-controlled variables

For these models the b_Genus_species coefficients are the species means at
log_Mass = 0 (the grand mean on the z-score scale). Because all species are
evaluated at the same mass value, size is held constant and comparisons
reflect shape differences only.

## 3b. Size-controlled variables

For these models we predict at each individual's own log_Mass, then average
draws within species. This marginalizes over the observed mass distribution
of each species rather than conditioning on a single reference value —
so the species-level posterior reflects what you'd expect for a typical
individual of that species at its actual size.
```{r}
ppred_size <- map_dfr(names(dist_mods_size), function(var_name) {
  
  mod <- dist_mods_size[[var_name]]
  if (is.null(mod)) return(NULL)
  
  tryCatch({
    mod %>%
      predicted_draws(
        newdata = dat_mice %>%
          select(Genus_species, catalog, log_Mass) %>%
          filter(!is.na(log_Mass))
      ) %>%
      as.data.frame() %>%
      group_by(Genus_species, .draw) %>%
      summarise(value = mean(.prediction, na.rm = TRUE), .groups = "drop") %>%
      mutate(var = var_name)
  }, error = function(e) {
    warning(glue("Prediction failed for {var_name}: {e$message}"))
    return(NULL)
  })
})

# Combine both sets
ppred_all <- bind_rows(post_no_size, ppred_size)
```


---

# 4. Plot

```{r}
# Variable order — preserve var_accurate ordering, restricted to modeled variables
var_acc_dist <- var_accurate[var_accurate %in% c(names(dist_mods_no_size),
                                                   names(dist_mods_size))]
n_shades <- floor(length(var_acc_dist) / 2)

shades <- data.frame(
  xmin = seq(1.5, by = 2, length.out = n_shades),
  xmax = seq(2.5, by = 2, length.out = n_shades),
  ymin = -Inf,
  ymax = Inf
)

# Raw data in long format for jittered points
dat_long <- dat_mice %>%
  select(Genus_species, catalog, all_of(var_acc_dist)) %>%
  pivot_longer(
    cols      = all_of(var_acc_dist),
    names_to  = "var",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  group_by(var) %>%
  mutate(value = scale(value)[,1]) %>%   # z-score within variable
  ungroup() %>%
  mutate(Genus_species = gsub("_", " ", as.character(Genus_species)))
```

```{r}
# Plot with raw jittered points overlaid
# Both stat_pointinterval and geom_point use the same dodge.width = 0.5
# to ensure species are aligned across layers
morpho_plot <- function(groups) {
  
  ppred_plot <- ppred_all %>%
    filter(Genus_species %in% groups) %>%
    mutate(Genus_species = gsub("_", " ", as.character(Genus_species)))
  
  raw_plot <- dat_long %>%
    filter(gsub(" ", "_", Genus_species) %in% groups)
  
  ppred_plot %>%
    ggplot(aes(
      x     = fct_relevel(var, var_acc_dist),
      y     = value,
      shape = Genus_species
    )) +
    geom_rect(
      inherit.aes = FALSE,
      data        = shades,
      mapping     = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      alpha       = 0.125
    ) +
    stat_pointinterval(
      .width    = 0.8,
      color     = "grey15",
      linewidth = 1.7,
      position  = position_dodge(width = 0.5)
    ) +
    geom_point(
      data     = raw_plot,
      mapping  = aes(x = fct_relevel(var, var_acc_dist), y = value,
                     shape = Genus_species),
      position = position_jitterdodge(dodge.width = 0.5, jitter.width = 0.1,
                                      jitter.height = 0),
      size     = 1.5,
      alpha    = 0.5,
      color    = "grey30"
    ) +
    scale_shape_manual(values = c(1, 15, 17, 19, 8)) +
    labs(x = "", y = "Z-score") +
    theme_pander(gM = FALSE) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.7),
      legend.text  = element_text(face = "italic"),
      legend.title = element_blank()
    ) +
    guides(shape = guide_legend(override.aes = list(size = 4)))
}

# Clean version without raw points
morpho_plot_clean <- function(groups) {
  
  ppred_plot <- ppred_all %>%
    filter(Genus_species %in% groups) %>%
    mutate(Genus_species = gsub("_", " ", as.character(Genus_species)))
  
  raw_plot <- dat_long %>%
    filter(gsub(" ", "_", Genus_species) %in% groups)
  
  ppred_plot %>%
    ggplot(aes(
      x     = fct_relevel(var, var_acc_dist),
      y     = value,
      shape = Genus_species
    )) +
    geom_rect(
      inherit.aes = FALSE,
      data        = shades,
      mapping     = aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      alpha       = 0.125
    ) +
    stat_pointinterval(
      .width    = 0.8,
      color     = "grey15",
      linewidth = 1.7,
      position  = position_dodge(width = 0.5)
    ) +
    scale_shape_manual(values = c(1, 15, 17, 19, 8)) +
    labs(x = "", y = "Z-score") +
    theme_pander(gM = FALSE) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      panel.border = element_rect(colour = "grey40", fill = NA, linewidth = 0.7),
      legend.text  = element_text(face = "italic"),
      legend.title = element_blank()
    ) +
    guides(shape = guide_legend(override.aes = list(size = 4)))
}
```

```{r}
morpho_plot(mice)
ggsave(here("Plots", "Morpho_Distributions.pdf"), height = 8, width = 13)

morpho_plot_clean(mice)
ggsave(here("Plots", "Morpho_Distributions_no_points.pdf"), height = 8, width = 13)
```
